Red Wine Exploration by Shantanu Kumar
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(ggplot2)
require(MASS)
require(gridExtra)
require(dplyr)
require(memisc)
require(GGally)
require(reshape2)
```

```{r echo=FALSE,message=FALSE,warning=FALSE}

Wine <- read.csv('wineQualityReds.csv')
str(Wine)
summary(Wine)
Wine$quality <- as.ordered(Wine$quality)

```

In this exercise, I will explore a data set on wine quality. 
The objective is to explore which chemical properties influence the
quality of red wines. I'll start by exploring the data using the statistical
program, R. As interesting relationships in the data are discovered, 
I'll produce and refine plots to illustrate them. 

There are 1599 observations of 13 numeric variables.
X appears to be the unique identifier.
Quality is an ordered, categorical, discrete variable. From the literature,
this was on a 0-10 scale, and was rated by at least 3 wine experts. 
The values ranged only from 3 to 8, with a mean of 5.64 and median of 6.
All other variables seem to be continuous quantities;
(w/ the exception of the .sulfur.dioxide suffixes).
From the variable descriptions, it appears that 
fixed.acidity ~ volatile.acidity and free.sulfur.dioxide ~ total.sulfur.dioxide
may possible by dependent, subsets of each other.

```{r,message=FALSE,warning=FALSE}
Wine$X <- as.factor(Wine$X)
table(Wine$quality)
```


# Univariate Plots Section

```{r echo=FALSE,message=FALSE,warning=FALSE}
grid.arrange(qplot(Wine$fixed.acidity),
             qplot(Wine$volatile.acidity),
             qplot(Wine$citric.acid),
             qplot(Wine$residual.sugar),
             qplot(Wine$chlorides),
             qplot(Wine$free.sulfur.dioxide),
             qplot(Wine$total.sulfur.dioxide),
             qplot(Wine$density),
             qplot(Wine$pH),
             qplot(Wine$sulphates),
             qplot(Wine$alcohol),
             qplot(Wine$quality),
             ncol = 4)
```

# Univariate Analysis

Looking at Wine Quality, we find that it forms a normal distribution, with most
of the data structured around wine quality of 5 or 6.

To explore better, we can create a new variable Rating, which classifies Wines
into Good, Bad, and Average depending on their quality scores.

BAD is Quality less than 5.
AVERAGE is Quality less than 7 but greater than or equal to 5.
GOOD is any number above and including 7.

```{r,message=FALSE,warning=FALSE}
Wine$rating <- ifelse(Wine$quality < 5, 'bad', ifelse(
  Wine$quality < 7, 'average', 'good'))
Wine$rating <- ordered(Wine$rating,
                     levels = c('bad', 'average', 'good'))
summary(Wine$rating)
```

It appears that density and pH are normally distributed, with a few outliers.
Fixed and volatile acidity, sulfur dioxides, sulphates, and alcohol seem to be
long-tailed.
Qualitatively, residual sugar and chlorides have extreme outliers.
Citric acid appeared to have a large number of zero values. This might be a 
case of non reporting.

```{r,message=FALSE,warning=FALSE}
ggplot(Wine,aes(x=fixed.acidity))+geom_histogram(fill='red')+scale_x_log10(breaks=4:15)+
  xlab('Fixed Acidity')

ggplot(Wine) + geom_histogram(aes(x=volatile.acidity),fill='blue')+
  scale_x_log10(breaks=seq(0.1,1,0.1))

ggplot(Wine) + 
  geom_histogram(aes(x=citric.acid),fill='green') +
  scale_x_log10()
```

As we could clearly see, citric acid was one feature that was found to be not 
normally distributed on a logarithmic scale. The transformation caused 132 
data points to be in the infinite range, telling us that 132 values are
0, since we know Log(0) is infinity.

Let's verify that.

```{r,message=FALSE,warning=FALSE}
print(length(subset(Wine, citric.acid == 0)$citric.acid))
print(log(0))
```

```{r,message=FALSE,warning=FALSE}

p1 <- ggplot(data = Wine, aes(x = residual.sugar)) +
  geom_histogram() +
  scale_x_continuous(lim = c(0, quantile(Wine$residual.sugar, 0.95))) +
  xlab('residual.sugar, 95th percentile truncated')

p2 <- p1 + scale_x_log10(breaks=1:9) + xlab('residual.sugar, log10')
grid.arrange(p1, p2, ncol=1)


p3 <- ggplot(data = Wine, aes(x = chlorides)) +
  geom_histogram() +
  scale_x_continuous(lim = c(0, quantile(Wine$chlorides, 0.95))) +
  xlab('chlorides, 95th percentile truncated')

p4 <- p3 + scale_x_log10(breaks=seq(0.01,0.11,0.02)) + xlab('chlorides, log10')
grid.arrange(p3, p4, ncol=1)

p5 <- ggplot(data = Wine, aes(x = sulphates)) +
  geom_histogram() +
  scale_x_continuous(lim = c(0, quantile(Wine$sulphates, 0.95))) +
  xlab('sulphates, 95th percentile truncated')

p6 <- p5 + scale_x_log10(breaks=seq(0,1,0.2)) + xlab('sulphates, log10')
grid.arrange(p5, p6, ncol=1)

rm(p1,p2,p3,p4,p5,p6)
```

## What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

While exploring the univariate histogram distributions, there did not appear to
be any bimodal or multimodal distributions that would warrant sub-classification
into categorical variables. I considered potentially splitting residual.sugar
into 'sweet wine' and 'dry wine', but a residual sugar of greater than 45 g/L
or g/m^3 is needed to classify as a sweet wine.

## Did you create any new variables from existing variables in the dataset?

I instantiated an ordered factor rating, classifying each wine sample as
'bad', 'average', or 'good'.

Upon further examination of the data set documentation, it appears that 
fixed.acidity and volatile.acidity are different types of acids; 
tartaric acid and acetic acid. I decided to create a combined variable, 
TAC.acidity, containing the sum of tartaric, acetic, and citric acid.

```{r,message=FALSE,warning=FALSE}
Wine$TAC.acidity <- Wine$fixed.acidity + Wine$volatile.acidity + 
Wine$citric.acid
qplot(Wine$TAC.acidity,main = 'Histogram of TAC Acidity (Tartaric+Acetic+Citric)')
```

## Boxplots are better suited in visualizing the outliers.

```{r,message=FALSE,warning=FALSE}
get_simple_boxplot <- function(column, ylab) {
  return(qplot(data = Wine, x = 'simple',
               y = column, geom = 'boxplot',
               xlab = '',
               ylab = ylab))
}

grid.arrange(get_simple_boxplot(Wine$fixed.acidity, 'fixed acidity'),
             get_simple_boxplot(Wine$volatile.acidity, 'volatile acidity'),
             get_simple_boxplot(Wine$citric.acid, 'citric acid'),
             get_simple_boxplot(Wine$TAC.acidity, 'TAC acidity'),
             get_simple_boxplot(Wine$residual.sugar, 'residual sugar'),
             get_simple_boxplot(Wine$chlorides, 'chlorides'),
             get_simple_boxplot(Wine$free.sulfur.dioxide, 'free sulf. dioxide'),
             get_simple_boxplot(Wine$total.sulfur.dioxide, 'total sulf. dioxide'),
             get_simple_boxplot(Wine$density, 'density'),
             get_simple_boxplot(Wine$pH, 'pH'),
             get_simple_boxplot(Wine$sulphates, 'sulphates'),
             get_simple_boxplot(Wine$alcohol, 'alcohol'),
             ncol = 4)
```

## Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

In univariate analysis, I chose not to tidy or adjust any data,
short of plotting a select few on logarithmic scales. Bivariate boxplots,
with X as rating or quality, will be more interesting in showing trends
with wine quality.


# Bivariate Plots Section

```{r echo=FALSE,message=FALSE,warning=FALSE}
set.seed(1)
Wine_sample <- Wine[,-which(names(Wine) %in% c('X', 'rating'))][sample(1:length(Wine$quality), 40), ]

get_bivariate_boxplot <- function(x, y, ylab) {
  return(qplot(data = Wine, x = x, y = y, geom = 'boxplot', ylab = ylab))
}

grid.arrange(get_bivariate_boxplot(Wine$quality, Wine$fixed.acidity,
                                   'fixed acidity'),
             get_bivariate_boxplot(Wine$quality, Wine$volatile.acidity,
                                   'volatile acidity'),
             get_bivariate_boxplot(Wine$quality, Wine$citric.acid,
                                   'citric acid'),
             get_bivariate_boxplot(Wine$quality, Wine$TAC.acidity,
                                   'TAC acidity'),
             get_bivariate_boxplot(Wine$quality, log10(Wine$residual.sugar),
                                   'residual sugar'),
             get_bivariate_boxplot(Wine$quality, log10(Wine$chlorides),
                                   'chlorides'),
             get_bivariate_boxplot(Wine$quality, Wine$free.sulfur.dioxide,
                                   'free sulf. dioxide'),
             get_bivariate_boxplot(Wine$quality, Wine$total.sulfur.dioxide,
                                   'total sulf. dioxide'),
             get_bivariate_boxplot(Wine$quality, Wine$density,
                                   'density'),
             get_bivariate_boxplot(Wine$quality, Wine$pH,
                                   'pH'),
             get_bivariate_boxplot(Wine$quality, log10(Wine$sulphates),
                                   'sulphates'),
             get_bivariate_boxplot(Wine$quality, Wine$alcohol,
                                   'alcohol'),
             ncol = 4)

grid.arrange(get_bivariate_boxplot(Wine$rating, Wine$fixed.acidity,
                                   'fixed acidity'),
             get_bivariate_boxplot(Wine$rating, Wine$volatile.acidity,
                                   'volatile acidity'),
             get_bivariate_boxplot(Wine$rating, Wine$citric.acid,
                                   'citric acid'),
             get_bivariate_boxplot(Wine$rating, Wine$TAC.acidity,
                                   'TAC acidity'),
             get_bivariate_boxplot(Wine$rating, log10(Wine$residual.sugar),
                                   'residual sugar'),
             get_bivariate_boxplot(Wine$rating, log10(Wine$chlorides),
                                   'chlorides'),
             get_bivariate_boxplot(Wine$rating, Wine$free.sulfur.dioxide,
                                   'free sulf. dioxide'),
             get_bivariate_boxplot(Wine$rating, Wine$total.sulfur.dioxide,
                                   'total sulf. dioxide'),
             get_bivariate_boxplot(Wine$rating, Wine$density,
                                   'density'),
             get_bivariate_boxplot(Wine$rating, Wine$pH,
                                   'pH'),
             get_bivariate_boxplot(Wine$rating, log10(Wine$sulphates),
                                   'sulphates'),
             get_bivariate_boxplot(Wine$rating, Wine$alcohol,
                                   'alcohol'),
             ncol = 4)
```

# Bivariate Analysis

From exploring these plots, it seems that a 'good' wine generally has 
these trends:

higher fixed acidity (tartaric acid) and citric acid,
lower volatile acidity (acetic acid)
lower pH (i.e. more acidic)
higher sulphates
higher alcohol
to a lesser extent, lower chlorides and lower density
Residual sugar and sulfur dioxides did not seem to have a dramatic impact on
the quality or rating of the wines. Interestingly, it appears that different
types of acid affect wine quality different; as such, TAC.acidity saw an 
attenuated trend, as the presence of volatile (acetic) acid accompanied
decreased quality.

By utilizing cor.test, I calculated the correlation for each of these
variables against quality:

```{r,message=FALSE,warning=FALSE}
simple_cor_test <- function(x, y) {
  return(cor.test(x, as.numeric(y))$estimate)
}

correlations <- c(
  simple_cor_test(Wine$fixed.acidity, Wine$quality),
  simple_cor_test(Wine$volatile.acidity, Wine$quality),
  simple_cor_test(Wine$citric.acid, Wine$quality),
  simple_cor_test(Wine$TAC.acidity, Wine$quality),
  simple_cor_test(log10(Wine$residual.sugar), Wine$quality),
  simple_cor_test(log10(Wine$chlorides), Wine$quality),
  simple_cor_test(Wine$free.sulfur.dioxide, Wine$quality),
  simple_cor_test(Wine$total.sulfur.dioxide, Wine$quality),
  simple_cor_test(Wine$density, Wine$quality),
  simple_cor_test(Wine$pH, Wine$quality),
  simple_cor_test(log10(Wine$sulphates), Wine$quality),
  simple_cor_test(Wine$alcohol, Wine$quality))
names(correlations) <- c('fixed.acidity', 'volatile.acidity', 'citric.acid',
                         'TAC.acidity', 'log10.residual.sugar',
                         'log10.chlordies', 'free.sulfur.dioxide',
                         'total.sulfur.dioxide', 'density', 'pH',
                         'log10.sulphates', 'alcohol')
correlations
```

As we see, the top 4 come to be:
alcohol
sulphates (log10)
volatile acidity
citric acid

Let's plot them against each other using RATING as a color and facet.

```{r,message=FALSE,warning=FALSE}
ggplot(data = Wine, aes(x = log10(sulphates), y = alcohol)) +
  facet_wrap(~rating) +
  geom_jitter(alpha=0.2) 

ggplot(data = Wine, aes(x = volatile.acidity, y = alcohol)) +
  facet_wrap(~rating) +
  geom_point(alpha=0.2)

ggplot(data = Wine, aes(x = citric.acid, y = alcohol)) +
  facet_wrap(~rating) +
  geom_point(alpha=0.2)

ggplot(data = Wine, aes(x = volatile.acidity, y = log10(sulphates))) +
  facet_wrap(~rating) +
  geom_jitter(alpha=0.2)

ggplot(data = Wine, aes(x = citric.acid, y = log10(sulphates))) +
  facet_wrap(~rating) +
  geom_point(alpha=0.2)

ggplot(data = Wine, aes(x = citric.acid, y = volatile.acidity)) +
  facet_wrap(~rating) +
  geom_point(alpha=0.2)
```

The relative value of these scatterplots are suspect; if anything, 
it illustrates how heavily alcohol content affects rating.
The weakest bivariate relationship appeared to be alcohol vs. citric acid.
The plots were nearly uniformly-distributed. The strongest relationship
appeared to be volatile acididty vs. citric acid, 
which had a negative correlation.

Examining the acidity variables, I saw strong correlations between them:

```{r,message=FALSE,warning=FALSE}
ggplot(data = Wine, aes(x = fixed.acidity, y = citric.acid)) +
  geom_point(alpha=0.3)
cor.test(Wine$fixed.acidity, Wine$citric.acid)

ggplot(data = Wine, aes(x = volatile.acidity, y = citric.acid)) +
  geom_point(alpha=0.3)
cor.test(Wine$volatile.acidity, Wine$citric.acid)

ggplot(data = Wine, aes(x = log10(TAC.acidity), y = pH)) +
  geom_point(alpha=0.3)
cor.test(log10(Wine$TAC.acidity), Wine$pH)
```

Most notably, base 10 logarithm TAC.acidity correlated very well with pH.
This is certainly expected, as pH is essentially a measure of acidity.
An interesting question to pose, using basic chemistry knowledge, is to ask 
what other components other than the measured acids are affecting pH. 
We can quantify this difference by building a predictive linear model, 
to predict pH based off of TAC.acidity and
capture the % difference as a new variable.

```{r,message=FALSE,warning=FALSE}
m <- lm(I(pH) ~ I(log10(TAC.acidity)), data = Wine)
Wine$pH.predictions <- predict(m, Wine)

# (observed - expected) / expected
Wine$pH.error <- (Wine$pH.predictions - Wine$pH)/Wine$pH

ggplot(data = Wine, aes(x = quality, y = pH.error)) +
  geom_boxplot(outlier.colour = 'red')
```

We can also add something interesting to our model, to check its accuracy.
The RMS Error.

```{r,message=FALSE,warning=FALSE}
rmse <- function(error)
{
    sqrt(mean(error^2))
}
rmse(m$residuals)
#Now, we train a Support Vector Machine.
require(e1071)
SVM <- svm(I(pH) ~ I(log10(TAC.acidity)), data = Wine)
Wine$pH.Predict.SVM <- predict(SVM,Wine)
Wine$pH.error.SVM <- (Wine$pH.Predict.SVM - Wine$pH)/Wine$pH

ggplot(data = Wine, aes(x = quality, y = pH.error.SVM)) +
  geom_boxplot(outlier.colour = 'red')

rmse(SVM$residuals)


```

We see that a SVM functions slightly better than a LM.

The first RMS is of the LM, the second of the SVM.

The median % error hovered at or near zero for most wine qualities.
Notably, wines rated with a quality of 3 had large negative error.
We can interpret this finding by saying that for many of the 'bad' wines, 
total acidity from tartaric, acetic, and citric acids were a worse predictor
of pH. Simply put, it is likely that there were other components--possibly
impurities--that changed and affected the pH.

As annotated previously, I hypothesized that free.sulfur.dioxide and 
total.sulfur.dioxide were dependent on each other. Plotting this:

```{r,message=FALSE,warning=FALSE}
ggplot(data = Wine, aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide)) +
  geom_point(alpha=0.2) +
  geom_smooth(se=F)

cor.test(Wine$free.sulfur.dioxide, Wine$total.sulfur.dioxide)
```

It is clear that there is a very strong relationship between the two.
Aside from TAC.acidity, this seemed to be the strongest bivariate relationship.
Additionally, despite the telling name descriptions, the clear 'floor' on this
graph hints that free.sulfur.dioxide is a subset of total.sulfur.dioxide.

# Multivariate Plots Section

```{r echo=FALSE,message=FALSE,warning=FALSE}
ggplot(data = Wine,
       aes(x = citric.acid, y = volatile.acidity,
           color = quality)) +
  geom_point() +
  facet_wrap(~rating)

ggplot(data = Wine,
       aes(x = alcohol, y = log10(sulphates),
           color = quality)) +
  geom_point() +
  facet_wrap(~rating)

ggplot(data = Wine,
       aes(x = pH, y = alcohol, color = quality)) +
  geom_point() +
  facet_wrap(~rating)

ggplot(data = Wine,
       aes(x = pH, y = alcohol, color = rating)) +
  geom_point() +
  facet_wrap(~quality)

```

# Multivariate Analysis

I primarily examined the 4 features which showed high correlation with quality.
These scatterplots were a bit crowded, so I faceted by rating AND by quality in
the final plot to illustrate clearly and a little more about the population
differences between good wines, average wines, and bad wines.
It's clear that a higher citric acid and lower volatile (acetic) acid
contributes towards better wines. Likewise, better wines tended to have
higher sulphates and alcohol content. Surprisingly, pH had very little visual
impact on wine quality, and was shadowed by the larger impact of alcohol.
Interestingly, this shows that what makes a good wine depends on the type
of acids that are present.

------

# Final Plots and Summary

### Plot One: Effects of Acid on Wine Quality
```{r echo=FALSE,message=FALSE,warning=FALSE}
grid.arrange(ggplot(data = Wine, aes(x = quality, y = fixed.acidity,
                                   fill = quality)) + 
               ylab('Fixed Acidity (g/dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=F),
             ggplot(data = Wine, aes(x = quality, y = volatile.acidity,
                                   fill = quality)) +
               ylab('Volatile Acidity (g/dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=F), 
             ggplot(data = Wine, aes(x = quality, y = citric.acid,
                                   fill = quality)) +
               ylab('Citric Acid (g/dm^3)') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=F), 
             ggplot(data = Wine, aes(x = quality, y = pH,
                                   fill = quality)) +
               ylab('pH') +
               xlab('Quality') +
               geom_boxplot()+guides(fill=F),top='Effects of Acid on Wine Quality')
?grid.arrange

```

### Description One
These subplots were created to demonstrate the effect of acidity and pH on 
wine quality. Generally, higher acidity (or lower pH) is seen in highly-rated
wines. To caveat this, a presence of volatile (acetic) acid negatively affected
wine quality. Citric acidity had a high correlation with wine quality, while
fixed (tartaric) acid had a smaller impact.

### Plot Two: Effect of Alcohol on Wine Quality
```{r echo=FALSE,message=FALSE,warning=FALSE}
ggplot(data = Wine, aes(x = quality, y = alcohol,
                      fill = rating)) +
  geom_boxplot(outlier.color = 'red') +
  ggtitle('Alcohol Levels in Different Wine Qualities') +
  xlab('Quality') +
  ylab('Alcohol (% volume)')
```

### Description Two
These boxplots demonstrate the effect of alcohol content on wine quality.
Generally, higher alcohol content correlated with higher wine quality.
However, as the outliers and intervals show, alchol content alone did not
produce a higher quality.

### Plot Three: What makes good wines, good, and bad wines, bad?
```{r echo=FALSE,message=FALSE,warning=FALSE}
ggplot(data = subset(Wine, rating != 'average'),
       aes(x = volatile.acidity, y = alcohol,
                      color = rating)) +
  geom_point(alpha=0.4) +
  ggtitle('Alcohol vs. Volatile Acidity and Wine Quality') +
  xlab('Volatile Acidity (g / dm^3)') +
  ylab('Alcohol (% volume)') + geom_smooth(method = 'lm',se=F)

```

### Description Three
This is perhaps the most telling graph. I subsetted the data to remove the
'average' wines, or any wine with a rating of 5 or 6. As the correlation tests
show, wine quality was affected most strongly by alcohol and volatile acidity.
While the boundaries are not as clear cut or modal, it's apparent that high
volatile acidity--with few exceptions--kept wine quality down. 
A combination of high alcohol content and low volatile
acidity produced better wines.


# Reflection

Through this exploratory data analysis, I was able to identify the key factors
that determine and drive wine quality, mainly: alcohol content, sulphates, 
and acidity. It is important to note, however, that wine quality is ultimately
a subjective measure, albeit measured by wine experts. That said, the 
correlations for these variables are within reasonable bounds. The graphs 
adequately illustrate the factors that make good wines 'good' and
bad wines 'bad'. Further study with inferential statistics could be done to 
quantitatively confirm these assertions.


